CrypTool Script \\ Mathematics and Cryptography

### 2003-03-10 script/en/chapters/numbertheory.tex
no message

% ..........................................................................
% --------------------------------------------------------------------------
% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\newpage
%\hypertarget{Kap_3}
\section{Introduction to Elementary Number Theory with Examples}
(Bernhard Esslinger, besslinger@web.de, July 2001, Updates: Dec. 2001, June 2002, Feb. 2003)

@@

We now know that patterns from number theory can be found everywhere in nature. 
For example, the ratio of 
%laevorotary and dextrorotary 
rotating counterclockwise and rotating clockwise 
spirals in a sunflower is equal to two consecutive 
Fibonacci\index{Fibonacci} numbers
\footnote{The sequence of Fibonacci numbers $(a_i)_{i \in \mathbb{N}}$ is defined 
by the ``recursive'' rule $a_1 := a_2 := 1$ and for all numbers  $n=1,2,3,\cdots$ 
we define $a_{n+2} := a_{n+1}+a_n$.  
This historical sequence can be found in many interesting forms in nature 
(for example, see \cite[p. 290 ff]{Graham1994}\index{Graham 1994} or the 
website of \hyperlink{knott}{Ron Knott}\index{Knott, Ron}, which is devoted 
to Fibonacci\index{Fibonacci} numbers). A lot is known about the Fibonacci 
sequence and it is used today as an important tool in mathematics.}, for 
example $21 : 34$.

@@

For this purpose we will use both theory and examples to explain how to perform calculations in finite sets and 
describe how these techniques are applied in cryptography. Particular attention will be paid to the traditional 
Diffie-Hellman (DH) and RSA public key procedures.\index{RSA}

@@

Cryptographically not only the uniqueness of the inverse is important, but also that the
set of possible values is exhausted.
\begin{theorem}\label{thm-zth-exhperm}
For $a,i\in \{1, \cdots, m-1 \}$ with ${\rm~gcd} (a,m)=1),$ then takes the product $a*i {\rm ~mod ~} m $ for a 
certain number $a$ all values from $\{1, \cdots ,m-1 \} $ (exhaustive permutation \index{Permutation} of the Length $m-1$)\footnote{See
also theorem \ref{thm-zth-ordp} in \hyperlink{Kap_3_9}{chapter 3.9 Multiplicative order and primitive roots}.}.
\end{theorem}
\vskip +10 pt
The following three examples\footnote{%
See \hyperlink{AppArith1}{Appendix D} for the source code to compute the tables using Mathematica\index{Mathematica}
or Pari-GP.\index{Pari-GP}
}
illustrate the properties of multiplicative inverses.

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\subsection{Euler function, Fermat's Little Theorem and Euler-Fermat}

@@

{\bf Example 1:} \\
The following table\footnote{%
See \hyperlink{AppArith3a}{Appendix D} for the source code to generate the table using Mathematica or Pari-GP.
} 
shows the values $a^i$ mod $11$ for the exponents $i = 1, 2, \cdots, 10$ and for the bases 
$a =  1, 2, \cdots, 10$ as well as the resulting value $ord_{11}(a)$ for each $a$:

@@

{\bf Example 2:} \\
The following table\footnote{%
See \hyperlink{AppArith3b}{Appendix D} for the source code to generate the table using Mathematica or Pari-GP.
} shows the values $a^i$ mod $45$ for the exponents $i = 1, 2, \cdots, 12$ and for the bases 
$a =  1, 2, \cdots, 12$ as well as the resulting value $ord_{45}(a)$ for each $a$:

@@

{\bf Example 4:} \\
The following table\footnote{%
See \hyperlink{AppArith3c}{Appendix D} for the source code to generate the table using Mathematica
or Pari-GP.
} answers the question as to whether the number $7$ is a primitive root of $46$.
The requirement/condition $gcd(7,46)=1$ is fulfilled.

@@

\hypertarget{RSABeweis} Using the Euler-Fermat theorem, we can ``prove'' the
RSA\footnote{The RSA procedure\index{RSA procedure} is the most common asymmetric 
cryptography procedure. Developed in 1978 by Ronald Rivest, Adi Shamir and Leonard Adleman, it can be used both for
signatures and for encryption. Cryptographers always associate this procedure with the abbreviation ``RSA''. However, 
please note that the abbreviation may be used in other fields to mean
something completely different.}$^,$\footnote{%
Within literature and movies not only classical methods are used (like the
secret message solved by Sherlock Holmes in ``The Adventure of the Dancing
Men'' by Arthur Conan  Doyle \index{Arthur Conan Doyle}), but also modern schemes. 
Examples are
\begin{itemize}
\item the movie ``House of Cards'', 1992, where autistic children
  communicate via primes,
\item the story ``The Dialogue of the Sisters'' by Dr.~C.~Elsner
  \index{Elsner 1999}, 1999
  (included in the CrypTool package\index{CrypTool} as PDF file), where the sisters
  communicate confidentially using a variant of RSA.
\end{itemize}
} 
procedure in the group $\mathbb{Z}_n^*$.

@@

%Therefor the fundamental characteristics of a special method are the first two points.
The following sections examine the RSA algorithm with respect to the first
two points.

@@

The best factorization method known today is a further development of the
General Number Field Sieve (GNFS), which was originally devised to
factor only numbers of a special form (like Fermat numbers).  The
complexity of solving the factorization problem with the GNFS is
asymptotically
$$
O(l) = e^{c \cdot (l \cdot \ln 2)^{1/3} \cdot  (\ln(l \cdot \ln(2))^{2/3} + o(l)}
$$
Please refer to: \\
\begin{list}{}{\setlength{\topsep}{-7 pt}}
  \item[] A. Lenstra, H. Lenstra:  
          {\em The development of the Number Field Sieve} 
          \cite{Lenstra1993}
  \item[] Robert D. Silverman:  
          {\em A Cost-Based Security Analysis of Symmetric and Asymmetric
	  Key Lengths} 
	  \cite{Silverman2000}
\end{list}
\vskip +12 pt

@@

Please refer to: {\em Handbook of Applied Cryptography} \cite{Menezes2001}

% --------------------------------------------------------------------------
\vskip +40pt
\subsubsection{Security Parameters because of progress in factorisation} \label{chptSecurityParam}

@@

      Please refer to:
      \begin{list}{}{\setlength{\topsep}{-7 pt}}
        \item[] C. Pomerance:  
                {\em The quadratic sieve factoring algorithm} 
                \cite{Pomerance1984}
      \end{list}
      \vskip +12pt

@@

      Please refer to:
      \begin{list}{}{\setlength{\topsep}{-7 pt}}
        \item[] J.P. Buhler, H.W. Lenstra, C. Pomerance:  
                {\em Factoring integers with the number field sieve} 
                \cite{Buhler1993}
      \end{list}
      \vskip +12 pt
\end{itemize}

@@

Within the last 20 years a lot of progress has been made. Estimations about
the future development of the ability to factor RSA modules vary and depend
on some assumptions:
\begin{itemize}
\item progression in computing performance\index{Performance} (Moore's law: every 18 month the
  computing power will double) and in grid computing 
\item development of new algorithms.
\end{itemize}
Within the last years the module bit length feasible for factorization
increased --- even without new algorithms --- by 10 bit per year. Larger
numbers require not only more time to be factored, but also huge RAM
storage for the solutions matrix being used by the best algorithms known
today.  This need for storage grows like the square root of the computation
time, i.\:e.\ also sub-exponentially. Because RAM availability increased
exponentially in the recent decades, it seems that this should not be the
limiting factor.

@@

% --------------------------------------------------------------------------
\vskip +40pt
\subsubsection{Further actual Research in Factorisation and Number Theory}
Prime numbers are part of many very topical research areas in number theory and computer science. 
Progress made with factorisation is bigger than estimated 5 years ago -- this
is not only founded on faster computers but also in new knowledge.

@@

% --------------------------------------------------------------------------
\vskip +20pt
\paragraph{Bernstein's paper and it's implication on the
  security of the RSA algorithm}
\label{RSABernstein} \index{Factorisation problem}\mbox{}
% \\
In his paper ``Circuits for integer factorization: a proposal'' 
(\href{http://cr.yp.to/djb.html}{\texttt{http://cr.yp.to/djb.html}}),
published November 2001, D.~J.\ Bernstein \cite{Bernstein2001} 
addresses the problem of
factorizing large integers. Therefore, his results are of relevance from a
RSA point of view.  As a main result Bernstein claims that the
implementation of the General Number Field Sieve algorithm (GNFS) can be
improved to factor, with the same effort as before, integers with three
times more digits.

@@

% --------------------------------------------------------------------------
\vskip +20pt
\paragraph{The TWIRL Device} \label{TWIRLDevice} \index{TWIRL Device}\mbox{}
% \\
In January 2003 Adi Shamir and Eran Tromer from the Weizmann Institute of Science published a preliminary draft called {\em ``Factoring Large Numbers with the TWIRL Device''} raising concerns about the security of key sizes till 1024 bits \cite{Shamir2003}.

@@

% --------------------------------------------------------------------------
\vskip +20pt
\paragraph{``Primes in P'': Primality testing is polynomial} \label{TWIRLDevice} \index{TWIRL Device}\mbox{}
% \\
In August 2002 the three Indian researchers M. Agrawal, N. Kayal and N. Saxena published the paper {\em ``PRIMES in P''} about a new primality testing algorithm 
\cite{Agrawal2002}. 
They discovered a polynomial time deterministic algorithm for determining if a number is prime or not.

@@

Current research results on this topic can be found at:\\
\href{http://www.mersenne.org/}{\texttt{http://www.mersenne.org/}}  and \\
\href{http://fatphil.org/maths/AKS/}{\texttt{http://fatphil.org/maths/AKS/}} .
\\

\vspace{4ex}
\begin{center}
\fbox{\parbox{15cm}{
    \emph{Hermann Hesse\footnotemark:}\\
    To let the possible happen, you again and again have to try the impossible.
}}
\end{center}
\addtocounter{footnote}{0}\footnotetext{%
  German/Swiss writer and Nobel Prize winner, July 2, 1877 $-$ August 9, 1962.}

% --------------------------------------------------------------------------
\vspace{4ex}
\subsubsection{Status regarding factorisation of concrete large numbers}
\label{NoteFactorisation}

@@

The 6 smaller prime factors, already found before have been:
$$
\begin{array}{c}
3, 1907, 425796183929, \\
1624700279478894385598779655842584377, \\
3802306738549441324432139091271828121 \quad {\rm and} \\
128064886830166671444802576129115872060027.
\end{array}
$$
The first 3 factors can be easily computed (e.g. using CrypTool\index{CrypTool} 
via menu {\bf Indiv. Procedures \textbackslash RSA Demonstration \textbackslash 
Factorisation of a Number}).
The next three prime factors were found by P.~Zimmerman
(\href{http://www.loria.fr/~zimmerma/ecmnet}{\tt http://www.loria.fr/\~{}zimmerma/ecmnet}), 
T.~Grandlund (\href{http://www.swox.se/gmp/}{\tt http://www.swox.se/gmp/})
and R. Harley during the years 1999 and 2000 using the elliptic curve factorisation method.

@@

% --------------------------------------------------------------------------
\subsubsection{The Diffie-Hellman key exchange protocol}
\index{Key agreement (key exchange)!Diffie-Hellman}

Whitfield Diffie, Martin E. Hellman and Ralph Merkle developed this DH key 
exchange protocol in Stanford in 1976
\footnote{%
With CrypTool\index{CrypTool} v1.3.04 you can execute the single steps of this 
exchange protocol with concrete numbers: \\
using menu {\bf Indiv. Procedures \textbackslash Diffie-Hellman Demonstration}.
}
.

@@

Before applying the RSA procedure to a text, we will first demonstrate it directly
using a number
\footnote{Using CrypTool\index{CrypTool} v1.3 you can solve this with
the ``Indiv.Procedures \textbackslash RSA Demonstration \textbackslash RSA Cryptosystem''.}.
\begin{itemize}
\item[{\bf 1.}] Let the selected prime numbers be $p=5$ and $q=11$. \\
Thus, $n=55$ and $J(n)=(p-1)*(q-1)=40$.
\item[{\bf 2.}] $e = 7$ (should lie between $11$ and $40$ and must be relatively prime to $40$).
\item[{\bf 3.}] $d = 23$ (since $23*7 \equiv 161 \equiv 1 {\rm ~(mod~} 40)$),
    \begin{itemize}
    \item[] $\rightarrow$ Public key of the recipient:  $(55, 7),$
    \item[] $\rightarrow$ Private key of the recipient: $(55, 23).$
    \end{itemize}
\item[{\bf 4.}] Let the message be the number $M = 2$ (so no division into blocks is required).
\item[{\bf 5.}] Encryption: $C \equiv 2^7 \equiv 18 {\rm ~(mod~}55).$
\item[{\bf 6.}] The cipher text is simply the number $C = 18$ (we therefore do not need to divide it into blocks).
\item[{\bf 7.}] Decryption: 
        $M \equiv 18^{23} \equiv 18^{(1+2+4+16)} \equiv 18*49*36*26 \equiv 2 {\rm ~(mod~}55).$
\end{itemize}

@@

We have the text ``ATTACK AT DAWN'' and the characters are coded in the following simple manner\footnote{Using
CrypTool\index{CrypTool} v1.3 you can solve this with the ``Indiv.Procedures \textbackslash RSA Demonstration \textbackslash RSA Cryptosystem''.
This is also described in the tutorial/scenario in CrypTool's online help [Options, specify alphabet, number system, block length\index{Block length} 2 and decimal representation].}:

@@

\label{SrcArith4a}
All 7 parts are encrypted using: $C \equiv M^{37}{\rm ~(mod~}3,713)$\footnote{%
See \hyperlink{AppArith4a}{Appendix D} for source code to do RSA
  encryption using Mathematica or Pari-GP.  You can also encrypt
  the message with CrypTool\index{CrypTool} v1.3 ``Indiv. Procedures \textbackslash RSA Demonstration \textbackslash Factorisation of a Number''.}: \\
{\tt 1404 2932 3536 0001 3284 2280 2235}

@@

The idea for this task\footnote{Using CrypTool\index{CrypTool} v1.3 you can solve this with the 
``Indiv.Procedures \textbackslash RSA Demonstration \textbackslash RSA Cryptosystem''.} is taken from the example in \cite[p. 215]{Eckert2003}\index{Eckert 2003}.

@@

We will work through the example in 2 variants. The steps 1 to 3 are common for both.\par
{\bf Key generation (steps 1 to 3)}:\\ \label{SrcArith4b}
{\bf 1.} $p=503, q=509 \quad (n= 256,027; \; J(n)=(p-1)(q-1)=255,016=2^3*127*251)$\footnote{%
See \hyperlink{AppArith4b}{Appendix D} for the source code to factorise the number $J(n)$ using Mathematica
or Pari-GP.
Using CrypTool\index{CrypTool} v1.3 you can solve this with the ``Indiv.Procedures \textbackslash RSA Demonstration \textbackslash Factorisation of a Number''.}. \\
{\bf 2.} $e=256,027$ (should lie between $509$ and $255,016$ and must be relatively prime to $255,016$)\footnote{$e$ cannot,
therefore, be $2, 127$ or $251$ ($65537 = 2^{16}+1$).\\
In real life, $J(n)$ is not factorised but rather the Euclidean algorithm 
is used for the selected e to guarantee that ${\rm gcd}(d,J(n))=1$.}.\\
{\bf 3.} $d=231,953$ \\\strut\quad\ (since $e \equiv d^{-1}{\rm ~mod~}J(n): ~65,537*231,953 \equiv 15,201,503,761 \equiv 1
{\rm ~(mod~}67,000)$)\footnote{Other possible combinations of $(e,d)$ include: $(3, 170,011)$, $(5, 204,013)$, $(7, 36,431)$.}.

@@

The letters are not combined\footnote{For secure procedures we need large numbers that assume -- as far as possible -- all 
values up to $n-1$. If the possible value set for the numbers in the message is too small, even large prime numbers 
cannot make the procedure secure.  An ASCII character is represented by $8$ bits. If we want larger values we must 
combine several numbers. Two characters need $16$ bits, whereby the maximum value that can be represented is $65536$. 
The modulus $n$ must then be greater than $2^{16} = 65536$. This is applied in variant 2.
When the numbers are combined, the leading zeros are kept in binary notation (just as if we were to write all numbers with $3$ 
digits in decimal notation above and were then to obtain the sequence 
{\tt 082 083,  065 032, 119 111,  114 107,  115 033)}.}!

\label{SrcArith4c}
Each character is encrypted using: $C = M^{65,537} {\rm ~(mod~} 256,027)$\footnote{%
See \hyperlink{AppArith4c}{Appendix D} for the source code for RSA exponentiation using Mathematica
or Pari-GP.
}: \\
{\tt
\begin{tabular}{lllll}
212984 & 025546 & 104529 & 031692 & 248407 \\
100412 & 054196 & 100184 & 058179 & 227433\\
\end{tabular} }

@@

Blocks are formed\footnote{\vskip +3 pt \tt \begin{tabular}{lll}
& binary representation  & decimal representation\\
01010010, 82 & 01010010 01010011 & =21075 \\
01010011, 83 & \\
01000001, 65 & 01000001 00100000 & =16672  \\
00100000, 32  \\
01110111, 119 & 01110111 01101111 & =30575 \\
01101111, 111 \\ 
01110010, 114 & 01110010 01101011 & =29291 \\
01101011, 107 \\
01110011, 115 & 01110011 00100001 & =29473 \\
00100001, 33: 
\end{tabular}} (each ASCII character is encoded into a 8 digit binary number below):\\
{\tt 21075 16672 30575 29291 29473}\footnote{Using CrypTool\index{CrypTool} v1.3 you can solve this with the ``Indiv.Procedures \textbackslash RSA
Demonstration \textbackslash RSA Cryptosystem'' 
with the following options: all 256 ASCII characters, b-adic, block length\index{Block length} 2 and decimal representation.}

>>>>

% ..........................................................................
% --------------------------------------------------------------------------
% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
%                 E l e m e n t a r e  Z a h l e n t h e o r i e
% /~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

\newpage
%\hypertarget{Kap_3}
\section{Introduction to Elementary Number Theory with Examples}
(Bernhard Esslinger, besslinger@web.de, July 2001, Updates: Dec. 2001, June 2002, Feb. 2003) \\

@@

We now know that patterns from number theory can be found everywhere in nature. 
For example, the ratio of 
%laevorotary and dextrorotary 
rotating counterclockwise and rotating clockwise 
spirals in a sunflower is equal to two consecutive 
Fibonacci\index{Fibonacci} numbers \footnote{%
The sequence of Fibonacci numbers $(a_i)_{i \in \mathbb{N}}$ is defined 
by the ``recursive'' rule $a_1 := a_2 := 1$ and for all numbers  $n=1,2,3,\cdots$ 
we define $a_{n+2} := a_{n+1}+a_n$.  
This historical sequence can be found in many interesting forms in nature 
(for example, see \cite[p. 290 ff]{Graham1994}\index{Graham 1994} or the 
website of \hyperlink{knott}{Ron Knott}\index{Knott, Ron}, which is devoted 
to Fibonacci\index{Fibonacci} numbers). A lot is known about the Fibonacci 
sequence and it is used today as an important tool in mathematics.}, for 
example $21 : 34$.

@@

For this purpose we will use both theory and examples to explain how 
to perform calculations in finite sets and describe how these techniques
are applied in cryptography. Particular attention will be paid to the 
traditional Diffie-Hellman\index{Diffie-Hellman} (DH) and RSA public 
key procedures.\index{RSA}

@@

Cryptographically not only the uniqueness of the inverse is important, but also that the
set of possible values is exhausted.
\begin{theorem}\label{thm-zth-exhperm}
For $a,i\in \{1, \cdots, m-1 \}$ with ${\rm~gcd} (a,m)=1),$ then takes the product $a*i {\rm ~mod ~} m $ for a 
certain number $a$ all values from $\{1, \cdots ,m-1 \} $ (exhaustive permutation \index{Permutation} of the Length $m-1$)\footnote{See
also theorem \ref{thm-zth-ordp} in \hyperlink{Kap_3_9}{chapter 3.9 Multiplicative order and primitive roots}.}.
\end{theorem}
\vskip +10 pt
The following three examples\footnote{%
See \hyperlink{AppArith1}{Appendix D} for the source code to compute
the tables using Mathematica\index{Mathematica} and Pari-GP.\index{Pari-GP}
}
illustrate the properties of multiplicative inverses.

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\subsection{Euler function, Fermat's little theorem and Euler-Fermat}

@@

{\bf Example 1:} \\
The following table\footnote{%
See \hyperlink{AppArith3a}{Appendix D} for the source code to generate
the table using Mathematica and Pari-GP.
} 
shows the values $a^i$ mod $11$ for the exponents $i = 1, 2, \cdots, 10$ and for the bases 
$a =  1, 2, \cdots, 10$ as well as the resulting value $ord_{11}(a)$ for each $a$:

@@

{\bf Example 2:} \\
The following table\footnote{%
See \hyperlink{AppArith3b}{Appendix D} for the source code to generate
the table using Mathematica and Pari-GP.
} shows the values $a^i$ mod $45$ for the exponents $i = 1, 2, \cdots, 12$ and for the bases 
$a =  1, 2, \cdots, 12$ as well as the resulting value $ord_{45}(a)$ for each $a$:

@@

{\bf Example 4:} \\
The following table\footnote{%
See \hyperlink{AppArith3c}{Appendix D} for the source code to generate
the table using Mathematica and Pari-GP.
} answers the question as to whether the number $7$ is a primitive root of $46$.
The requirement/condition $gcd(7,46)=1$ is fulfilled.

@@

\hypertarget{RSABeweis} Using the Euler-Fermat theorem, we can ``prove'' the
RSA\footnote{The RSA procedure\index{RSA procedure} is the most common asymmetric 
cryptography procedure. Developed in 1978 by Ronald Rivest, Adi Shamir and Leonard Adleman, it can be used both for
signatures and for encryption. Cryptographers always associate this procedure with the abbreviation ``RSA''. However, 
please note that the abbreviation may be used in other fields to mean
something completely different.}$^,$\footnote{%
Within literature and movies not only classical methods are used (like the
secret message solved by Sherlock Holmes in ``The Adventure of the Dancing
Men'' by Arthur Conan  Doyle \index{Arthur Conan Doyle}), but also modern schemes. 
Examples are
% hier die items nicht einrücken ! xxxxxxxxxxxxxxx
\begin{itemize}
\item the movie ``House of Cards'', 1992, where autistic children
  communicate via primes,
\item the story ``The Dialogue of the Sisters'' by Dr.~C.~Elsner
  \index{Elsner 1999}, 1999
  (included in the CrypTool package\index{CrypTool} as PDF file), where the sisters
  communicate confidentially using a variant of RSA.
\end{itemize}
} 
procedure in the group $\mathbb{Z}_n^*$.

@@

%Therefore the fundamental characteristics of a special method are the first two points.
The following sections examine the RSA algorithm with respect to the first
two points.

@@

The best factorization method known today is a further development of the
General Number Field Sieve (GNFS), which was originally devised to
factor only numbers of a special form (like Fermat numbers).  The
complexity of solving the factorization problem with the GNFS is
asymptotically
$$
O(l) = e^{c \cdot (l \cdot \ln 2)^{1/3} \cdot  (\ln(l \cdot \ln(2))^{2/3} + o(l)}
$$
Please refer to: \\
\begin{list}{}{\setlength{\topsep}{-7 pt}}
  \item[] A. Lenstra, H. Lenstra:  
          {\em The development of the Number Field Sieve} 
          \cite{Lenstra1993}.
  \item[] Robert D. Silverman:  
          {\em A Cost-Based Security Analysis of Symmetric and Asymmetric
	  Key Lengths} 
	  \cite{Silverman2000}.
\end{list}
\vskip +12pt

@@

Please refer to: {\em Handbook of Applied Cryptography} \cite{Menezes2001}.

% --------------------------------------------------------------------------
\vskip +40pt
\subsubsection{Security parameters because of progress in 
               factorisation} \label{chptSecurityParam}

@@

      Please refer to:
      \begin{list}{}{\setlength{\topsep}{-7 pt}}
        \item[] C. Pomerance:  
                {\em The quadratic sieve factoring algorithm} 
                \cite{Pomerance1984}.
      \end{list}
      \vskip +12pt

@@

      Please refer to:
      \begin{list}{}{\setlength{\topsep}{-7 pt}}
        \item[] J.P. Buhler, H.W. Lenstra, C. Pomerance:  
                {\em Factoring integers with the number field sieve} 
                \cite{Buhler1993}.
      \end{list}
      \vskip +12pt
\end{itemize}

@@

Within the last 20 years a lot of progress has been made. Estimations about
the future development of the ability to factor RSA modules vary and depend
on some assumptions:
\begin{itemize}
  \item progression in computing performance\index{Performance} (Moore's 
      law\index{Moore's law}: every 18 month the computing power will double) 
      and in grid computing\index{grid computing}.
  \item development of new algorithms.
\end{itemize}
Within the last years the module bit length feasible for factorization
increased --- even without new algorithms --- by 10 bit per year. Larger
numbers require not only more time to be factored, but also huge RAM
storage for the solutions matrix being used by the best algorithms known
today.  This need for storage grows like the square root of the computation
time, i.\:e.\ also sub-exponentially. Because RAM availability increased
exponentially in the recent decades, it seems that this should not be the
limiting factor.

@@

% --------------------------------------------------------------------------
\vskip +40pt
\subsubsection{Further actual research in factorisation and number theory}
Prime numbers are part of many very topical research areas in number theory and computer science. 
Progress made with factorisation is bigger than estimated 5 years ago -- this
is not only founded on faster computers but also in new knowledge.

@@

% --------------------------------------------------------------------------
\vskip +20pt
\paragraph{Bernstein's paper and it's implication on the
  security of the RSA algorithm}
\label{RSABernstein} \index{Factorisation problem}\mbox{}

In his paper ``Circuits for integer factorization: a proposal'' 
(\href{http://cr.yp.to/djb.html}{\texttt{http://cr.yp.to/djb.html}}),
published November 2001, D.~J.\ Bernstein \cite{Bernstein2001} 
addresses the problem of
factorizing large integers. Therefore, his results are of relevance from a
RSA point of view.  As a main result Bernstein claims that the
implementation of the General Number Field Sieve algorithm (GNFS) can be
improved to factor, with the same effort as before, integers with three
times more digits.

@@

% --------------------------------------------------------------------------
\vskip +20pt
\paragraph{The TWIRL Device} \label{TWIRLDevice} \index{TWIRL Device}\mbox{}

In January 2003 Adi Shamir and Eran Tromer from the Weizmann Institute of Science published a preliminary draft called {\em ``Factoring Large Numbers with the TWIRL Device''} raising concerns about the security of key sizes till 1024 bits \cite{Shamir2003}.

@@

% --------------------------------------------------------------------------
\vskip +20pt
\paragraph{``Primes in P'': Primality testing is polynomial} \label{TWIRLDevice} \index{TWIRL Device}\mbox{}

In August 2002 the three Indian researchers M. Agrawal, N. Kayal and N. Saxena published the paper {\em ``PRIMES in P''} about a new primality testing algorithm 
\cite{Agrawal2002}. 
They discovered a polynomial time deterministic algorithm for determining if a number is prime or not.

@@

Current research results on this topic can be found at:\\
\begin{itemize}
  \item[] \href{http://www.mersenne.org/}
               {\texttt{http://www.mersenne.org/}} \\
          \href{http://fatphil.org/maths/AKS/}
               {\texttt{http://fatphil.org/maths/AKS/}} 
\end{itemize}
\vskip +10 pt

\vspace{4ex}
\begin{center}
\fbox{\parbox{15cm}{
    \emph{Hermann Hesse\footnotemark:}\\
    To let the possible happen, you again and again have to try the impossible.
}}
\end{center}
\addtocounter{footnote}{0}\footnotetext{%
  German/Swiss writer and Nobel Prize winner, July 2, 1877 $-$ August 9, 1962.}
% --------------------------------------------------------------------------
\vspace{4ex}
\subsubsection{Status regarding factorisation of concrete large numbers}
\label{NoteFactorisation}

@@

The 6 smaller prime factors, already found before have been:
$$
\begin{array}{c}
3, 1907, 425796183929, \\
1624700279478894385598779655842584377, \\
3802306738549441324432139091271828121 \quad {\rm and} \\
128064886830166671444802576129115872060027.
\end{array}
$$
The first 3 factors can be easily computed \footnote{%
e.g. using CrypTool\index{CrypTool} via menu 
{\bf Indiv. Procedures \textbackslash{} RSA Demonstration \textbackslash{} 
Factorisation of a Number}.
}
The next three prime factors were found by P.~Zimmerman
(\href{http://www.loria.fr/~zimmerma/ecmnet}{\tt http://www.loria.fr/\~{}zimmerma/ecmnet}), 
T.~Grandlund (\href{http://www.swox.se/gmp/}{\tt http://www.swox.se/gmp/})
and R. Harley during the years 1999 and 2000 using the elliptic curve factorisation method.

@@

% --------------------------------------------------------------------------
\subsubsection{The Diffie-Hellman key exchange protocol}
\index{Diffie Whitfield} 
\index{Hellman Martin} 
\index{Key agreement (key exchange)!Diffie-Hellman}
\index{Diffie-Hellman}

Whitfield Diffie, Martin E. Hellman and Ralph Merkle developed this DH key 
exchange protocol in Stanford in 1976\footnote{%
With CrypTool\index{CrypTool} v1.3.04  this exchange protocol has been
visualized: you can execute the single steps with concrete numbers using 
menu {\bf Indiv. Procedures \textbackslash{} Diffie-Hellman Demonstration}.
}
.

@@

Before applying the RSA procedure to a text, we will first demonstrate it directly
using a number
\footnote{Using CrypTool\index{CrypTool} v1.3 you can solve this with
the {\bf Indiv.Procedures \textbackslash{} RSA Demonstration \textbackslash{} RSA Cryptosystem}.}.
\begin{itemize}
\item[{\bf 1.}] Let the selected prime numbers be $p=5$ and $q=11$. \\
Thus, $n=55$ and $J(n)=(p-1)*(q-1)=40$.
\item[{\bf 2.}] $e = 7$ (should lie between $11$ and $40$ and must be relatively prime to $40$).
\item[{\bf 3.}] $d = 23$ (since $23*7 \equiv 161 \equiv 1 {\rm ~(mod~} 40)$),
    \begin{itemize}
    \item[] $\rightarrow$ Public key of the recipient:  $(55, 7),$
    \item[] $\rightarrow$ Private key of the recipient: $(55, 23).$
    \end{itemize}
\item[{\bf 4.}] Let the message be the number $M = 2$ (so no division into blocks is required).
\item[{\bf 5.}] Encryption: $C \equiv 2^7 \equiv 18 {\rm ~(mod~}55).$
\item[{\bf 6.}] The cipher text is simply the number $C = 18$ (we therefore do not need to divide it into blocks).
\item[{\bf 7.}] Decryption: 
        $M \equiv 18^{23} \equiv 18^{(1+2+4+16)} \equiv 18*49*36*26 \equiv 2 {\rm ~(mod~}55).$
\end{itemize}

@@

We have the text ``ATTACK AT DAWN'' and the characters are coded in the following simple manner\footnote{Using
CrypTool\index{CrypTool} v1.3 you can solve this with the {\bf Indiv.Procedures \textbackslash{} RSA Demonstration \textbackslash{} RSA Cryptosystem}.
This is also described in the tutorial/scenario in CrypTool's online help [Options, specify alphabet, number system, block length\index{Block length} 2 and decimal representation].}:

@@

\label{SrcArith4a}
All 7 parts are encrypted using: $C \equiv M^{37}{\rm ~(mod~}3,713)$\footnote{%
See \hyperlink{AppArith4a}{Appendix D} for source code to do RSA
  encryption using Mathematica and Pari-GP.  You can also encrypt
  the message with CrypTool\index{CrypTool} v1.3 {\bf Indiv. Procedures \textbackslash{} RSA Demonstration \textbackslash{} Factorisation of a Number}.}: \\
{\tt 1404 2932 3536 0001 3284 2280 2235}

@@

The idea for this task\footnote{Using CrypTool\index{CrypTool} v1.3 you can 
solve this with the {\bf Indiv.Procedures \textbackslash{} RSA Demonstration 
\textbackslash{} RSA Cryptosystem}.} is taken from the example in 
\cite[p. 271]{Eckert2003}\index{Eckert 2003}.

@@

We will work through the example in 2 variants. The steps 1 to 3 are common for both.\par
{\bf Key generation (steps 1 to 3)}:\\ \label{SrcArith4b}
{\bf 1.} $p=503, q=509 \quad (n= 256,027; \; J(n)=(p-1)(q-1)=255,016=2^3*127*251)$\footnote{%
See \hyperlink{AppArith4b}{Appendix D} for the source code to factorise
the number $J(n)$ using Mathematica and Pari-GP.
Using CrypTool\index{CrypTool} v1.3 you can solve this with the 
{\bf Indiv.Procedures \textbackslash{} RSA Demonstration \textbackslash{} 
Factorisation of a Number}.}. \\
{\bf 2.} $e=256,027$ (should lie between $509$ and $255,016$ and must be relatively prime to $255,016$)\footnote{$e$ cannot,
therefore, be $2, 127$ or $251$ ($65537 = 2^{16}+1$).\\
In real life, $J(n)$ is not factorised but rather the Euclidean algorithm 
is used for the selected e to guarantee that ${\rm gcd}(d,J(n))=1$.}.\\
{\bf 3.} $d=231,953$ \\\strut\quad\ (since $e \equiv d^{-1}{\rm ~mod~}J(n): ~65,537*231,953 \equiv 15,201,503,761 \equiv 1
{\rm ~(mod~}67,000)$)\footnote{Other possible combinations of $(e,d)$ include: $(3, 170,011)$, $(5, 204,013)$, $(7, 36,431)$.}.

@@

The letters are not combined\footnote{For secure procedures we need large numbers that assume -- as far as possible -- all 
values up to $n-1$. If the possible value set for the numbers in the message is too small, even large prime numbers 
cannot make the procedure secure.  An ASCII character is represented by $8$ bits. If we want larger values we must 
combine several numbers. Two characters need $16$ bits, whereby the maximum value that can be represented is $65536$. 
The modulus $n$ must then be greater than $2^{16} = 65536$. This is applied in variant 2.
When the numbers are combined, the leading zeros are kept in binary notation (just as if we were to write all numbers with $3$ 
digits in decimal notation above and were then to obtain the sequence 
{\tt 082 083,  065 032, 119 111,  114 107,  115 033}).}!

\label{SrcArith4c}
Each character is encrypted using: $C = M^{65,537} {\rm ~(mod~} 256,027)$\footnote{%
See \hyperlink{AppArith4c}{Appendix D} for the source code for RSA 
exponentiation using Mathematica and Pari-GP.
}: \\
{\tt
\begin{tabular}{lllll}
212984 & 025546 & 104529 & 031692 & 248407 \\
100412 & 054196 & 100184 & 058179 & 227433\\
\end{tabular} }

@@

Blocks are formed\footnote{\vskip +3 pt \tt \begin{tabular}{lll}
& binary representation  & decimal representation\\
01010010, 82 & 01010010 01010011 & =21075 \\
01010011, 83 & \\
01000001, 65 & 01000001 00100000 & =16672  \\
00100000, 32  \\
01110111, 119 & 01110111 01101111 & =30575 \\
01101111, 111 \\ 
01110010, 114 & 01110010 01101011 & =29291 \\
01101011, 107 \\
01110011, 115 & 01110011 00100001 & =29473 \\
00100001, 33: 
\end{tabular}} (each ASCII character is encoded into a 8 digit binary number below):\\
{\tt 21075 16672 30575 29291 29473}\footnote{Using CrypTool\index{CrypTool} v1.3 you can solve this with the {\bf Indiv.Procedures \textbackslash{} RSA
Demonstration \textbackslash{} RSA Cryptosystem} 
with the following options: all 256 ASCII characters, b-adic, block length\index{Block length} 2 and decimal representation.}
